Writing Trash Bin

### 2024-07-24 paper.qmd
reorganised "visualising uncertainy as signal" section

# Visualising uncertainty as signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations. Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all. It is acting as signal. This is easiest to understand with an example.

@@

Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. This group are *also* viewing uncertainty as a signal, albeit more explicitly than the decision making group. If the purpose of uncertainty in these cases is to convey information rather than obfuscate it, then there is no reason to assume these "uncertainty visualisations" would behave any differently to normal information visualisations. This is exactly what the literature shows.

Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat. Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease. This is the logic that results in predictable results. @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. @Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. These experiments all have the same fundamental finding, if we create a problem by obfuscating information, then the representation that makes the obfuscated information easier to extract outperforms those that don't. This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks. We have known perceptual tasks have more noise than text representations for a long time [@Cleveland1984] and it has little to do with uncertainty.

@@

This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012]. We belive this disagreement comes from two different conceptualisations of what an uncertainty  visualisation *is*. This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.

@@

@@

# Including uncertainty as an indepdendent variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.

@@

### There is no such thing as a test for everything
As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise. Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.

@@

# Evaluating uncertainty visualisations
# Current methods
# Testing signal supression
## Current methods
### Secondary effects

@@

### Direct questions

@@

@@

# What is an uncertainty visualisation?
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

The authors of these taxonomies did not intend for them to be complete descriptions of uncertainty visualisations, however they do paint a picture of the what authors consider to be important. There are common themes in the taxonomies that ensure we differentiate between a sample and a mass, consider if uncertainty should be depicted with the estimate or as a second variable, identify the parallel of hypothesis testing, and consider the dimensionality, sample size and precision depicted by the visualisation. These taxonomies may highlight perceived important differences, but they do not help us understand what *is* or *is not* an uncertainty visualisation. For that we need to understand *why* uncertainty visualisation exists as a sub-field at all.

### What is uncertainty?

The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

#### Uncertainty as a by-product of inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

## What is an uncertainty visualisation?
The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication. Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.

# How do we evaluate uncertainty visualisations?
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

>>>>

# Visualising uncertainty as signal
Uncertainty visualisation is not only motivated by signal-suppression, and we would be remiss to ignore these alternative motivations. In this section we discuss alternative motivations for uncertainty visualization, and explain why they are not the focus of this review.

## Uncertainty as a variable of interest
Other authors do not describe uncertainty as important for decision making, but rather explicitly state it as variable of importance in of itself [@Blenkinsop2000]. While uncertainty can provide useful information in decision making, it is important to recognize the "uncertainty" in these cases is not acting as "uncertainty" at all. It is acting as signal. This is easiest to understand with an example.

Extracting information using a graphic requires a perceptual task, so it will always be less accurate than explicitly reading the value provided in text form [@Cleveland1984].

This is why text representations or graphics that allow users to simply count a value to extract the required statistic outperform those that require any perceptual tasks.

## Uncertainty for decision making
Some authors claim the purpose of uncertainty is to improve decision making [@Ibrekk1987; @uncertchap2022; @Hullman2016; @Cheong2016; @Boone2018; @Padilla2017]. This group are *also* viewing uncertainty as a signal, although it is less obvious in this case.

@@

This obfuscation problem translates to unanswerable questions in uncertainty evaluation experiments.

If we create a problem by obfuscating information, then the representation that makes the obfuscated information easiest to extract outperforms those that don't.

@@

@@

## What is an uncertainty visualisation?

This seems to be an accepted fact for all visualisations *except* for uncertainty visualisation, and this creates disagreement in what is even considered to be an uncertainty visualisation. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, and @Wickham2011 suggests their product plot framework, which includes histograms and bar charts, should be extended to include uncertainty however at least one or both of these charts are used in a significant number of uncertainty visualisation experiments [@Ibrekk1987; @Olston2002; @Zhao2023; @Hofmann2012].

We belive this disagreement comes from two different conceptualizations of what an uncertainty  visualisation *is*. This issue causing confusion in the literature was also commented on by @Kinkeldey2014, who explained it was unclear if uncertainty should be treated as a variable, as metadata, or as something else entirely [@Kinkeldey2014]. There are those that describe *any* visualisation of error as an "uncertainty visualisation" and those that believe "uncertainty visualisation" is a function that can be *applied* to any graphic that depicts an estimate. The former group view the purpose of uncertainty visualisation to provide signal about a distribution, while the later believe it should act as noise to obfuscate a signal.

@@

# Visualising uncertainty as a variable
Upon hearing that uncertainty needs to be included for transparency, the solutions may seem obvious. You may think "well, I will just add a dimension to my plot that includes uncertainty". This makes sense, as this is the simplest way to add uncertainty to an existing graphic is to simply map uncertainty to an unused visual channel.

@@

### Uncertainty and inference
The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

As we move through these methods, it seems that the validity of any overarching insight becomes more visible at the cost of our ability to extract particular values of signal or noise. Therefore, given that the primary goal of visualisation *is* insights [@North2006], visualisation authors should err on the side of representing suppressed signal as a single variable, rather than visualising uncertainty separately using two different channels.

The concept of uncertainty as a by-product of inference, not a latent feature of data, extends naturally to visualisation. 
There are two primary reasons for making a visualisation, to perform exploratory data analysis (EDA) or communication. Communication involves identifying a signal we want to communicate and designing the visualisation that best conveys that, while EDA involves creating a versatile visualisation without an explicit purpose and using it to extract several signals.

@@

# Evaluating uncertainty visualisations

## Uncertainty disappears if you stare directly at it

# Current methods

# Testing signal supression
If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

## Current methods
### Secondary effects

@@

### Direct questions
Several authors have connected the issues around defining uncertainty to inference, however it is often discussed as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017].

Experiments where users are asked to make a decision based on a probability, and experiments where users need to extract a variance or some feature of the distribution both have the implicit belief that uncertainty visualisation is a "class" of visualisation, and a "good" uncertainty visualisation allows us to extract all information with ease.

@@

Uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] with little consideration for how difficult the information is to extract from the chat. For example, @Ibrekk1987 found that participants were more accurate at extracting a statistic when it could be directly read off the graphic, than when it required an area estimate (which is the case if using the PDF), or when there was no visual indicator for the statistic at all (which is the case when using the CDF of an asymmetric function). @Hullman2015 found that a visualisation that allows viewers to count outcomes to estimate a probability outperformed one that required a complicated area calculation. @Hofman2020 and @Zhang2022 found that participants were better at answering questions about a prediction intervals when shown a prediction interval instead of a sampling distribution. @Gschwandtnei2016 found that graphics where the required statistic could be directly read off the plot outperformed those that involved guesswork due to a gradually decreasing line. @Cheong2016 found that participants made better decisions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map.

@@

# How do we evaluate uncertainty visualisations?

### 2024-07-24 writingdumpster.qmd
reorganised "visualising uncertainy as signal" section

# Removed Sections
Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014]. @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.

>>>>

# Removed Sections
@Kinkeldey2014 categorised uncertainty according to five criteria depicted in fig-tax which considers if a visualisation is implicit or explicit, intrinsic or extrinsic, visually integrable or seperable, coincident or adjacent, and static or dynamic. A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which seems to functionally align with the intrinsic/extrinsic distinction by @Kinkeldey2014. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", function that is used to describe a random variable, is used to describe both the data and the uncertainty for all dimensions. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution").

Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014]. @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed.
